Simple repression draft

### 2020-06-04 doc/appendix_01_nonbursty.tex
MRM pass through appendix 1 to fix spacing. Section is basically ready

Since mRNA production can only take place in the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transition that can happen for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ with m mRNAs  to the active state $A$
    with m mRNAs.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ with m mRNAs to the repressor bound state $R$ with
    m mRNAs.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions occur with probability $r\Delta t$. The same is true
for the promoter state transitions, each occurs with probability $k_R^\pm \Delta
t$. As for the mRNA degradation events, these transitions depend on the current number of
mRNA molecules since the more molecules of mRNA there are, the more will decay during
a given time interval. Each molecule has a constant probability $\gamma \Delta t$ of 
being degraded, so the total probability for an mRNA degradation event to happen
is computed by multiplying this probability by the current number of mRNAs.

To see these terms in action let us compute the probability of having $m$ mRNA
at time $t + \Delta t$ in the transcriptionally active state. This takes the 
form
\begin{equation}
\begin{aligned}
p_A(m, t + \Delta t) &= p_A(m, t) \\
&+ \overbrace{(r\Delta t) p_A(m-1, t)}^{m-1 \rightarrow m}
- \overbrace{(r\Delta t) p_A(m, t)}^{m \rightarrow m+1}\\
% \;\;\;\text{(mRNA production)} \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_A(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_A(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&+\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
-\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R},
% \;\;\;\text{(promoter state transition)}
\end{aligned}
\end{equation}
where the overbrace indicates the corresponding state transitions. Notice that
the second to last term on the right-hand side is multiplied by $p_R(m, t)$ 
since the transition from state $R$ to state $A$ depends on the probability of
being in state $R$ to begin with. It is through this term that the dynamics of the two probability
distribution functions ($p_R(m,t)$ and $p_A(m,t)$ are coupled. An equivalent equation can be written for
the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
bound state, the only difference being that the mRNA production rates are 
removed, and the sign for the promoter state transitions are inverted. This is
\begin{equation}
\begin{aligned}
p_R(m, t + \Delta t) &= p_R(m, t) \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_R(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_R(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&-\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
+\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R}.
\end{aligned}
\end{equation}

>>>>

Since mRNA production can only take place in the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transition that can happen for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ with $m$ mRNAs  to the
    active state $A$ with $m$ mRNAs.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ with $m$ mRNAs to the repressor
    bound state $R$ with $m$ mRNAs.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions occur with probability $r\Delta t$. The same is true for
the promoter state transitions, each occurs with probability $k_R^\pm \Delta t$.
As for the mRNA degradation events, these transitions depend on the current
number of mRNA molecules since the more molecules of mRNA there are, the more
will decay during a given time interval. Each molecule has a constant
probability $\gamma \Delta t$ of being degraded, so the total probability for an
mRNA degradation event to happen is computed by multiplying this probability by
the current number of mRNAs.

To see these terms in action let us compute the probability of having $m$ mRNA
at time $t + \Delta t$ in the transcriptionally active state. This takes the 
form
\begin{equation}
\begin{aligned}
p_A(m, t + \Delta t) &= p_A(m, t) \\
&+ \overbrace{(r\Delta t) p_A(m-1, t)}^{m-1 \rightarrow m}
- \overbrace{(r\Delta t) p_A(m, t)}^{m \rightarrow m+1}\\
% \;\;\;\text{(mRNA production)} \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_A(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_A(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&+\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
-\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R},
% \;\;\;\text{(promoter state transition)}
\end{aligned}
\end{equation}
where the overbrace indicates the corresponding state transitions. Notice that
the second to last term on the right-hand side is multiplied by $p_R(m, t)$
since the transition from state $R$ to state $A$ depends on the probability of
being in state $R$ to begin with. It is through this term that the dynamics of
the two probability distribution functions ($p_R(m,t)$ and $p_A(m,t)$) are
coupled. An equivalent equation can be written for the probability of having $m$
mRNA at time $t + \Delta t$ while in the repressor bound state, the only
difference being that the mRNA production rates are removed, and the sign for
the promoter state transitions are inverted. This is
\begin{equation}
\begin{aligned}
p_R(m, t + \Delta t) &= p_R(m, t) \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_R(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_R(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&-\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
+\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R}.
\end{aligned}
\end{equation}
